Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-02 methods.tex
adding to methods and detection

\section{Experimental Methods }

@@

\subsection{The Programs}
Include traces of full \col \svd and \gcc
\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.475\textwidth}
    \includegraphics[width=0.95\textwidth]{figs/colFullTS}
    \caption{\col Time Series}
    \label{fig:col-ts}
  \end{subfigure}%
  \begin{subfigure}[t]{0.475\textwidth}
    \includegraphics[width=0.95\textwidth]{figs/gccfullts}
    \caption{\gcc Time Series}
    \label{fig:gcc-ts}
  \end{subfigure}
  \caption{In (a) the instructions executed per CPU clock cycle
    (IPC) during the execution of \col. Each point is the average IPC during a 100,000
    instruction period. Similarly in (b) is a time series of the IPC during the execution of \gcc.}\label{fig:sample-ts}
\end{figure}

>>>>

\section{Experimental Methods and Time Series Explanations}

@@

\subsection{The Programs}
Include traces of full \col \svd and \gcc
\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.475\textwidth}
    \includegraphics[width=0.95\textwidth]{figs/colFullTS}
    \caption{\col Time Series}
    \label{fig:col-ts}
  \end{subfigure}%
  \begin{subfigure}[t]{0.475\textwidth}
    \includegraphics[width=0.95\textwidth]{figs/gccfullts}
    \caption{\gcc Time Series}
    \label{fig:gcc-ts}
  \end{subfigure}
    \begin{subfigure}[t]{0.475\textwidth}
    \includegraphics[width=0.95\textwidth]{figs/SVD1RegimesColored}
    \caption{\svd Time Series with Regime Coloring}
    \label{fig:svd-ts-colored}
  \end{subfigure}
  \caption{[[Maybe this should be just a picture of each time series and then have another figure with color after explaining heuristically the reigmes. In (a) the instructions executed per CPU clock cycle
    (IPC) during the execution of \col. Each point is the average IPC during a 100,000
    instruction period. Similarly in (b) is a time series of the IPC during the execution of \gcc.}\label{fig:sample-ts}
    \end{figure}

\svd brings up a very interesting point: systems change over time. While it is clear in Figure \ref{fig:sample-ts} (a) and (b) that a single system is being observed, it also appears \emph{visually} that the complexity of these two time series is consistent over time, i.e., while they are both complicated they don't appear to increase or decrease in complexity over the range of the program. \svd (seen in Figure \ref{fig:svd-ts-colored}) is different however, at least visually. Figure \ref{fig:svd-ts-colored} is a single trace of \svd from start to finish, however as time progresses it is clear that something drastically changes in the underlying generating process and the structure of the time series is visually very different from section to section. This is caused by the code of \svd moving between different subroutines. We call these changes in \svd dynamics  \emph{\svd regimes} and we have colored each of the six regimes different colors in Figure \ref{fig:svd-ts-colored}. The advantage to splitting \svd into these regimes is to explore how complexity and predictability evolve over time for a system in drift. In Section \ref{sec:wpeRegime} we explore and validate the choice of these visually selected regime windows extending techniques from \cite{cao2004det}. The purpose of this paper is not to rigorously explore regime detection but to explore quantifying complexity of a time series. As such, making this split simply serves as providing 90 unique time series\footnote{Six Regimes with 15 individual runs each.} to explore. For notational convience we refer to these signals as {\tt dgesdd$_i$} with $i \in \{1\dots6\}$ where $i$ corresponds to a regime of \svd. The regimes are labeled from left to right.

### 2014-03-02 wpe.tex
adding to methods and detection

The weighted permutation entropy for the {\tt SVD} program is given in
Fig.~\ref{fig:wwpe}. To generate this image a window of 5,000 values slid over
the time series. Within each of those windows, the statistics over words of
length 4 are computed and the WPE is calculated. The gray bands denote regions
where the 5,000 value window overlapped visually-distinct regimes. It can be
seen that the behaviors of the weighted permutation entropy vary between
regimes. [[I think here it would be good to add a paragraph explaining the windowed WPE was used for regime choices on SVD...emphasizing  that over a time series permutation entropy fluctuates illustrating within a single time series different levels of complexity and predictability exist. Maybe point at some of the predicting predictability papers.]]

>>>>

\subsection{Regime Selection:Detecting complexity shift in a time series}\label{sec:wpeRegime}

{\color{red} this is from a previous paper I wrote, if this is used we need to update the plots to use wpe and windowed wpe instead of standard pe, also not sure how much of this is necessary but added as starting point}

Permutation entropy is a measure of the complexity of time series as a whole, but it can also be used to calculate a dynamical shift in a time series as illustrated in \cite{cao2004det}. To detect dynamical changes in a time series we follow the methods of \cite{cao2004det}. Partition the time series  $\{x_t\}_{t = 1,\dots,T}$ into overlapping\footnote{Following \cite{cao2004det} we use maximal overlapping block. (A window shift of 1.)} (or nonoverlapping) blocks. Calculate $H(n)$ for each block, and treat $H(n)$ as a function of time.  According to \cite{cao2004det} drastic changes in permutation entropy can indicate change in the underlying model.

In addition to introducing this block method, \cite{cao2004det} introduce the use of a \emph{lag} when calculating permutation entropy. This is identical to the lag used in delay coordinate embedding and often the same language is used. Cao et al. describe the process of symbolization in \cite{bandt2002per} as embedding the time series in an $m$ dimensional symbol space. When we symbolize a scalar time series,$\{x_t\}_{t = 1,\dots,T}$, we first embed the scalar time series into $m$ dimensional vectors before matching this to a symbol in $\mathcal{S}_m$. So pre-symbolization our time series changes from scalar to objects of the form:  $$[x(i),\,x(i+1),\,\dots,x(i+(m-1)]$$ This vector is then mapped by $\phi$ to a symbol in $\mathcal{S}_m$. Cao et al. argue that we can view this as embedding the time series in an $m$ dimensional space with lag 1. Cao et al. then state ``Since in practice the optimal $L$[lag] may be different from 1, we shall present the idea for any $m$[embedding dimension] and $L$."\footnote{No further explanation is given about why this is the case.} To not confuse notation we define $\tau$ to be the lag parameter. The process of symbolization then starts with transforming the scalar time series into a set of vectors of the form $$[x(i),\,x(i+\tau),\,\dots,x(i+(m-1)\tau]$$ where $\tau\ge1$.

\subsubsection{Example: Concatenated Time Series}\label{sec:concatenate}
As an example of this detection, consider Figure \ref{fig:ts-banded} (a). This is a plot of a concatenated time series comprised of several orbits of the Logistic map with varying parameters as well as intermittent noise. Each concatenated band (either an orbit of the Logistic map or noise) is 50,000 points. The bands of noise are generated with Matlab's \verb|randsample|, using the chaotic bands as the sampling population. Each Logistic map trajectory begins at $x=0.6$. The period-two trajectories have an $R$ of 3.2, the period-three trajectory has an R of 3.838, and the chaotic trajectories have an $R$ of 3.65.

The concatenated time series is comprised of two period-two trajectories($t=0-50,000$ and $t=250,000-300,000$), one period-three trajectory($t=100,000-150,000$), two chaotic trajectories ($t=50,000-100,000$ and $t=200,000-250,000$, and two different bands of noise($t=150,000-200,000$ and $t=300,000-350,000$). Visually the bands of noise and the chaotic trajectory are indistinguishable. If this were an observed time series from a real process, it would be impossible to visually identify which bands were generated deterministically and which were simply noise. If we calculate block permutation entropy as discussed in \cite{cao2004det} and plot this as a function of time it becomes very clear which bands have structure and which do not.

In Figure \ref{fig:ts-banded} (b) we plot the concatenated time series, as well as permutation entropy as a function of time(red, blue and green curves). The red, blue and green curves are the permutation entropy calculated using $m=5$ and $\tau = 1,2,3$ (respectively). Using these curves it is clear that bands 2 and 5 ($t=50,000-100,000$ and $t=200,000-250,000$) are chaotic because they have structure\footnote{Permutation entropy less than one.}. On the other hand, bands 4 and 7 ($t=150,000-200,000$ and $t=300,000-350,000$) have no structure\footnote{Permutation entropy of one.} and are the noise bands. This is in fact the case when comparing to the way the time series was constructed.

The main advantage we can see to the use of lags other than 1 is for analysis of periodicity. This is also the major advantage listed in \cite{cao2004det}. Notice in the period 2 bands the blue curve (permutation entropy calculated with a lag of 2) is zero. This means that with a lag of 2 the period 2 orbit is perfectly structured\footnote{Equivalently, not enough permutations are being realized to impact the calculation of $H(n)$}. If we increase the lag to 3 (the green curve) the two period-two bands have a little bit of structure(as some permutations will be realized) but the period 3 band has zero permutation entropy.

%TSLogistic = [period2Logistic;chaosLogistic;period3Logistic;randomSampleFromMap2;chaosLogistic;period2Logistic;randomSampleFromMap3];

%\begin{figure}
%  \centering
%  \subfigure{\includegraphics[width=\columnwidth]{unused-figs/ts-banded-alone}}
%    {(a)}
%
%  \subfigure{\includegraphics[width=\columnwidth]{unused-figs/ts-banded-pe}}
%  {(b)}
%  \caption{ (a) The concatenated time series discussed in Section~\ref{sec:concatenate}. (b) Plot of the concatenated time series from Section~\ref{sec:concatenate}, as well as permutation entropy as a function of time(blue,green and red curves). Each permutation entropy curve shown is calculated using $m=5$, however, the red,blue and green curve use a $\tau = 1,2,3$ respectively.}\label{fig:ts-banded}
%\end{figure}

and what is it good for with forward pointer to the other sections

\subsubsection{Example: The transient Logistic map}
In Section~\ref{sec:concatenate} we discussed a time series which stayed in a parameter regime for long amounts of time and then abruptly changed. As we showed, permutation was a great proxy to detect this change. Another type of parameter change that may occur in a physical system is constant parameter drift, i.e., some parameter slowly changes over time. To explore this type of parameter drift we construct the same time series as was used in \cite{cao2004det}. Iterate the logistic map, starting with $R=2.8$, then at every iteration increase $R$ by $10^{-5}$ until $R=4$. The time series, which can be seen in Figure~\ref{fig:translogistic} appears very similar to the standard bifurcation diagram for the Logistic map. The curves in red and blue are the permutation entropy calculated with $m=5$ and $\tau = 1,2$ (respectively).

 As we can see, as bifurcations occur in the dynamics the permutation entropy changes as well. An interesting feature of block calculations are the spikes that occur at bifurcation points. As the block begins to encompass multiple regimes, e.g., fixed point and  period two, permutations are being realized for both regimes. This overlap causes an increase in the time series complexity, and thus a spike in the permutation entropy. The dips you see in the second half correspond to bands of periodicity which occur between the chaos. It is worth noting, that the blue curve (permutation entropy with lag 2) does not go to zero in the period 2 parameter range. This is because this is not a true period 2 orbit but transient that is moving across \emph{several} period 2 orbits, thus it maintains a higher level of complexity, albeit still very low.

 This illustrates that permutation entropy can not only distinguish between instant (and drastic) parameter shifts but also very slow and gradual parameter drift.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\columnwidth]{unused-figs/tstranslog-perm}
  \caption{Transient Logistic map time series and permutation entropy curves. Both permutation entropy curves are calculated with $m=5$, the red and blue correspond to $\tau = 1,2$ respectively.}
  \label{fig:translogistic}
\end{figure}
%\begin{enumerate}
%\item do with henon bifurcation. what happens near the broken chaotic window

%\end{enumerate}

The weighted permutation entropy for the \svd program is given in
Fig.~\ref{fig:wwpe}. To generate this image a window of 5,000 values slid over
the time series. Within each of those windows, the statistics over words of
length 4 are computed and the WPE is calculated. The gray bands denote regions
where the 5,000 value window overlapped visually-distinct regimes. It can be
seen that the behaviors of the weighted permutation entropy vary between
regimes. [[I think here it would be good to add a paragraph explaining the windowed WPE was used for regime choices on SVD...emphasizing  that over a time series permutation entropy fluctuates illustrating within a single time series different levels of complexity and predictability exist. Maybe point at some of the predicting predictability papers.]]
